Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-05-08 manuscript.tex
Work on introduction

@@

Lots of community efforts toward coupled transport-depletion calculations with
OpenMC~\citep{gul2017ane,lanversin2017icone,lanversin2019phd,liu2019nst,zhuang2020pne,zhao2020ned,zhang2020ane}.

When materials in a system are subject to irradiation over a long period of
time, nuclides within the material will transmute due to nuclear reactions as
well as spontaneous radioactive decay. The time-dependent process by which
nuclides transmute under irradiation is known as \emph{depletion} or
\emph{burnup}. To accurately analyze nuclear systems, it is often necessary to
predict how the composition of materials will change since this change results
in a corresponding change in the solution of the transport equation. The
equation that governs the transmutation and decay of nuclides inside of an
irradiated environment can be written as
\begin{equation}
  \frac{dN_i(t)}{dt} = \sum\limits_j
  \underbrace{\left [ \underbrace{f_{j \rightarrow i} \int_0^\infty dE \;
  \sigma_j (E, t) \phi(E,t)}_\text{transmutation} +
  \underbrace{\lambda_{j\rightarrow i}}_\text{decay} \right ]
  N_j(t)}_{\text{Production of nuclide }i\text{ from nuclide }j}
  - \underbrace{\left [\underbrace{\int_0^\infty dE \; \sigma_i
  (E,t) \phi(E,t)}_\text{transmutation} +
  \underbrace{\sum\limits_j \lambda_{i\rightarrow j}}_\text{decay} \right ]
  N_i(t)}_{\text{Loss of nuclide }i}
\end{equation}
where $N_i$ is the density of nuclide $i$ at time $t$, $\sigma_i$ is the
transmutation cross section for nuclide $i$ at energy $E$, $f_{j \rightarrow i}$
is the fraction of transmutation reactions in nuclide $j$ that produce nuclide
$i$, and $\lambda_{j \rightarrow i}$ is the decay constant for decay modes in
nuclide $j$ that produce nuclide $i$. Note that we have not included the spatial
dependence of the flux or cross sections. As one can see, the equation simply
states that the rate of change of $N_i$ is equal to the production rate minus
the loss rate. Because the equation for nuclide $i$ depends on the nuclide
density for possibly many other nuclides, we have a system of first-order
differential equations. To form a proper initial value problem, we also need the
nuclide densities at time 0:
\begin{equation}
    N_i(0) = N_{i,0}.
\end{equation}
These equations can be written more compactly in matrix notation as
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\mathbf{n}}{dt} = \mathbf{A}(\mathbf{n},t)\mathbf{n}, \quad \mathbf{n}(0) =
  \mathbf{n}_0
\end{equation}
where $\mathbf{n} \in \mathbb{R}^n$ is the nuclide density vector,
$\mathbf{A}(\mathbf{n},t) \in \mathbb{R}^{n\times n}$ is the burnup matrix
containing the decay and transmutation coefficients, and $\mathbf{n}_0$ is the
initial density vector. Note that the burnup matrix depends on $\mathbf{n}$
because the solution to the transport equation depends on the nuclide densities.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

>>>>

When a material in a system is subject to irradiation over a long period of
time, nuclides within the material will transmute due to nuclear reactions and
spontaneous radioactive decay. The time-dependent process by which nuclides
transmute under irradiation is known as \emph{depletion}, \emph{burnup}, or
\emph{transmutation}. Modeling this process plays an important role in the
design and licensing of nuclear reactors~\citep{betzler2019ned}. Consequently,
many codes have been developed that solve the underlying equations for the
change in material compositions over time.

When modeling nuclear reactors, depletion is inherently linked with solving the
neutron transport equation. Nuclear reaction rates, which determine the rate at
which nuclides are transmuted, must be provided from a transport code.
Conversely, the change in material composition directly affects the solution of
the transport equation. As will be shown in \cref{sec:methods}, this results in
a non-linear coupling between transport and depletion. The implication is that
depletion codes are rarely used in isolation---normally, they are directly
coupled with a neutron transport code.

With continued advances in computing, Monte Carlo (MC) particle transport
simulations are now widely used in reactor design and analysis in a variety of
roles. The earliest efforts to couple MC particle transport with depletion date
back to the 1990s~\citep{moore1995inel,trellue1998lanl}. These early efforts
involved taking existing MC and depletion codes, such as
MCNP~\citep{goorley2012nt} and ORIGEN~\citep{croff1983nt}, and putting a
``wrapper'' around them that orchestrated the transfer of solutions from one
code to the other. Thus, the actual transport/depletion codes themselves
required no source code changes. In the last decade, many MC code developers
have chosen to directly include a depletion solver as a first-rate feature,
including MCNPX~\citep{waters2007aip} (which has since been merged into
MCNP6~\citep{goorley2012nt}), Serpent~\citep{leppanen2015ane}, and
MC21~\citep{griesheimer2015ane}.

In the present work, we describe a new depletion solver in
OpenMC~\citep{romano2015ane1}, an open source, community-developed MC particle
transport code. To date, OpenMC has not had a built-in depletion solver. This
has led to numerous efforts in the community toward developing coupled
transport-depletion calculations with
OpenMC~\citep{gul2017ane,lanversin2017icone,lanversin2019phd,liu2019nst,zhuang2020pne,zhao2020ned,zhao2020cpc,zhang2020ane}.
While these efforts have proven the viability of performing such coupled
calculations with OpenMC, none of them has resulted in a validated,
openly-available, and generalized capability that OpenMC users can take
advantage of. The capability that is described herein is part of the OpenMC
project and can be found in version 0.11 of the code or later.

The paper is organized as follows. In \cref{sec:methods}, we describe the
methods and algorithms used to solve the depletion equations, the implementation
within OpenMC, and how nuclear data is used to create depletion chains. Then, in
\cref{sec:results}, we present results on two problems that demonstrate the
accuracy of OpenMC when compared to Serpent~\citep{leppanen2015ane}. In
\cref{sec:conclusions}, we summarize the findings, discuss limitations, and
provide commentary on future avenues of inquiry that may be promising.

The equation that governs the transmutation and decay of nuclides inside of an
irradiated environment can be written as
\begin{equation}
  \frac{dN_i(t)}{dt} = \sum\limits_j
  \underbrace{\left [ \underbrace{f_{j \rightarrow i} \int_0^\infty dE \;
  \sigma_j (E, t) \phi(E,t)}_\text{transmutation} +
  \underbrace{\lambda_{j\rightarrow i}}_\text{decay} \right ]
  N_j(t)}_{\text{Production of nuclide }i\text{ from nuclide }j}
  - \underbrace{\left [\underbrace{\int_0^\infty dE \; \sigma_i
  (E,t) \phi(E,t)}_\text{transmutation} +
  \underbrace{\sum\limits_j \lambda_{i\rightarrow j}}_\text{decay} \right ]
  N_i(t)}_{\text{Loss of nuclide }i}
\end{equation}
where $N_i$ is the density of nuclide $i$ at time $t$, $\sigma_i$ is the
transmutation cross section for nuclide $i$ at energy $E$ and time $t$, $\phi$
is the neutron flux at energy $E$ and time $t$, $f_{j \rightarrow i}$ is the
fraction of transmutation reactions in nuclide $j$ that produce nuclide $i$, and
$\lambda_{j \rightarrow i}$ is the decay constant for decay modes in nuclide $j$
that produce nuclide $i$. Note that we have not included the spatial dependence
of the flux or cross sections. As one can see, the equation simply states that
the rate of change of $N_i$ is equal to the production rate minus the loss rate.
Because the equation for nuclide $i$ depends on the nuclide density for possibly
many other nuclides, we have a system of first-order differential equations. To
form a proper initial value problem, we also need the nuclide densities at time
0:
\begin{equation}
    N_i(0) = N_{i,0}.
\end{equation}
These equations can be written more compactly in matrix notation as
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\mathbf{n}}{dt} = \mathbf{A}(\mathbf{n},t)\mathbf{n}, \quad \mathbf{n}(0) =
  \mathbf{n}_0
\end{equation}
where $\mathbf{n} \in \mathbb{R}^n$ is the nuclide density vector,
$\mathbf{A}(\mathbf{n},t) \in \mathbb{R}^{n\times n}$ is the burnup matrix
containing the decay and transmutation coefficients, and $\mathbf{n}_0$ is the
initial density vector. Note that the burnup matrix depends on $\mathbf{n}$
because the solution to the transport equation depends on the nuclide densities.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
